---
title: "featureCounts-DESeq2 gene level"
subtitle: ""
author: ""
date: "`r paste('Last update:', format(Sys.time(), '%B %d, %Y'))`"
output: 
  html_document:
    toc: true
    toc_float: 
      collapsed: false
    toc_depth: 3
    number_sections: true
    theme: spacelab
    highlight: default
---

# Description
**Muscle-secreted neurturin couples fiber oxidative metabolism and slow motor neuron identity**

Brief description of the dataset
	In this study we evaluate the effects of muscle-derived neurturin on skeletal muscle innervation, metabolism, and performance.

*GEO Dataset:* <a href="https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE155068" target="_blank" rel="noopener">GSE155068</a>

*Overall design:* RNA-sequencing analysis of gastrocnemius muscles from muscle-specific NRTN transgenic mice (with NRTN expression driven by the human alpha-skeletal actin promoter) and wild-type littermates.

*Reference:*

- Correia JC, Kelahmetoglu Y, Jannig PR, Schweingruber C, Shvaikovskaya D, Zhengye L, Cervenka I, Khan N, Stec M, Oliveira M, Nijssen J, Martínez-Redondo V, Ducommun S, Azzolini M, Lanner JT, Kleiner S, Hedlund E, Ruas JL. 2021. Muscle-secreted neurturin couples myofiber oxidative metabolism and slow motor neuron identity. Cell Metabolism 33:2215-2230.e8. <a href="https://doi.org/10.1016/j.cmet.2021.09.003" target="_blank" rel="noopener">Link</a>


# Setup environment

```{r, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# define variables ==========================================
filename_prefix <- "" # leave empty if not necessary
species <- "mouse" # mouse, human, pig, rat
ensembl_version <- 102 # define ensembl release version

file_formats <- list(
  #  "png",
  "pdf"
)

# load libraries
required_Packages_Install <- c(
  "tidyverse",
  "DT",
  "DESeq2",
  "vsn",
  "biomaRt",
  "RColorBrewer",
  "ggrepel",
  "pheatmap",
  "readxl",
  "writexl"
)

for (Package in required_Packages_Install) {
  if (!require(Package, character.only = TRUE)) {
    BiocManager::install(Package, dependencies = TRUE)
  }
  library(Package, character.only = TRUE)
}

# source custom functions
source("load_functions.R") # source custom functions
```

Defined parameters
```{r, echo = FALSE}
cat("Filename prefix:", filename_prefix, "\n")
cat("Species:", species, "\n")
cat("Ensembl version:", ensembl_version, "\n")
```

# Data analysis

## Metadata

Read sample info column names:
```{r}
colnames(read.csv("../02_metadata/sampleInfo.csv",
  header = TRUE
))
```

Create metadata.csv:
```{r}
countsFolder <- "../01_results/counts/featureCounts"

metadata <- read.csv("../02_metadata/sampleInfo.csv",
    header = TRUE)
```

save metadata
```{r}
filename <- "metadata"
write.csv(metadata, paste0(
  "../02_metadata/",
  filename_prefix,
  filename, ".csv"
),
row.names = F, quote = F
)

saveRDS(metadata, paste0(
  "../02_metadata/",
  filename_prefix,
  filename,
  ".Rds"
))
```

```{r echo=FALSE}
cat("Saved file:", paste0(
  "../02_metadata/",
  filename_prefix,
  filename, ".csv"
))
```

### Metadata table
```{r}
filename <- "metadata"
metadata %>% DT::datatable(
  extensions = "Buttons",
  options = list(
    scrollX = TRUE,
    pageLength = nrow(metadata),
    dom = "Blfrtip",
    buttons = list(
      list(extend = "csv", filename = paste0(filename_prefix, filename)),
      list(extend = "excel", filename = paste0(filename_prefix, filename))
    )
  )
)
```

## DESeq2
Full rank with paired subjects
object construction
```{r}
dds <- DESeqDataSetFromFeatureCounts(metadata,
  countsFolder,
  design = ~ group
)
dds <- estimateSizeFactors(dds)
nrow(dds)
table(rowSums(assay(dds, "counts")) == 0)
```

filter low counts
```{r}
# # Option 1
# dds <- dds[rowSums(counts(dds, normalized=TRUE)) >= 1, ] ## remove genes with zero normalized counts
# dim(dds)
#
# # Option 2
# dds <- dds[rowSums(counts(dds, normalized=TRUE) >= 10), ] ## remove genes with less than 10 normalized counts across all samples
# dim(dds)
#
# # Option 3
# dds <- dds[rowSums(counts(dds, normalized=TRUE) >= 5 ) >= 3] ## remove genes where less than 3 samples with normalized counts greater than or equal to 5.
# dim(dds)

# Option 4
smallestGroupSize <- metadata %>%
  dplyr::group_by(group) %>%
  dplyr::summarise(samples = dplyr::n()) %>%
  pull(samples) %>%
  min()
dds <- dds[rowSums(counts(dds, normalized = TRUE) >= 1) >= smallestGroupSize, ] ## remove genes where less than X samples (smallestGroupSize) with normalized counts greater than or equal to 1.
dim(dds)
```

Relevel
```{r}
levels(dds$group)
dds$group <- factor(dds$group,
  levels = c(
    "WT",
    "HSA-NRTN"
  )
)
levels(dds$group)
```

Standard analysis
```{r}
dds <- DESeq(dds)
```

save dds file
```{r}
filename <- "dds"
saveRDS(dds, file = paste0(
  "../01_results/deseq2/",
  filename_prefix,
  filename, ".Rds"
))
```


# Quality control plots

## PCA plot 
```{r, fig.width = 3, fig.height = 3, fig.align = "center"}
palette_pca <- c("#1F78B4", "#E31A1C", "#FDBF6F")

pca <- plot_pca_deseq(dds,
  palette = palette_pca,
  plot_center = T,
  group = "Group"
) +
  #coord_cartesian(xlim = c(-60, 30), expand = TRUE) +
  ggtitle("**Principal component analysis**")
pca
```
Save plot

```{r}
filename <- "PCA_plot"
for (file_format in file_formats) {
  ggsave_fixed(
    paste0(
      "../03_figures/",
      filename_prefix,
      filename, ".",
      file_format
    ),
    units = "mm",
    plot_width = 40,
    plot_height = 40,
    width = 60,
    height = 60
  )
}
```

## PCA plot with labels
```{r, fig.width = 3, fig.height = 3, fig.align = "center"}
pca +
  geom_text_repel(aes(label = name),
    size = 5 / ggplot2:::.pt,
    nudge_x = 1,
    nudge_y = 0.2,
    segment.colour = NA
  )
```

```{r}
# Save plot
filename <- "PCA_plot_labels"

for (file_format in file_formats) {
  ggsave_fixed(
    paste0(
      "../03_figures/",
      filename_prefix,
      filename, ".",
      file_format
    ),
    units = "mm",
    plot_width = 40,
    plot_height = 40,
    width = 60,
    height = 60
  )
}
```

## Sample-to-sample distances
```{r, fig.width = 6, fig.height = 6, fig.align = "center"}
vsd <- vst(dds, blind = FALSE)
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$Group, vsd$Patient, sep = " - ")
colnames(sampleDistMatrix) <- NULL
heatmap <- pheatmap::pheatmap(sampleDistMatrix,
  clustering_distance_rows = sampleDists,
  clustering_distance_cols = sampleDists,
  show_rownames = T, show_colnames = T,
  cellwidth = 6,
  cellheight = 6,
  treeheight_row = 10,
  treeheight_col = 10,
  fontsize = 6,
  border_color = NA,
  col = colorRampPalette(rev(brewer.pal(9, "Blues")))(255)
)
```

```{r}
# Save plot
filename <- "sample_distances"
for (file_format in file_formats) {
  ggsave(
    plot = heatmap,
    paste0(
      "../03_figures/",
      filename_prefix,
      filename, ".",
      file_format
    ),
    units = "mm",
    dpi = 300,
    width = 3 * length(metadata$sample),
    height = 3 * length(metadata$sample)
  )
}
```

# Normalized gene counts

## biomaRt annotation
retrieve gene ensembl annotation 
```{r}
annotation <- getBM(
  filters = "ensembl_gene_id",
  attributes = c(
    "ensembl_gene_id",
    symbol,
    "chromosome_name",
    "gene_biotype"
  ),
  values = rownames(dds),
  mart = useEnsembl("ensembl",
    dataset = dataset,
    # mirror = "useast",
    version = ensembl_version
  )
)
```

Save biomaRt annotation
```{r}
filename <- "biomart_annotation"
write.csv(annotation, paste0(
  "../01_results/deseq2/",
  filename_prefix,
  filename,
  ".csv"
),
row.names = F, quote = F
)

saveRDS(annotation, paste0(
  "../01_results/deseq2/",
  filename_prefix,
  filename,
  ".Rds"
))
```

## DESeq2 Normalized counts
Create data frame for normalize counts
```{r}
normalized_counts <- annotation %>%
  inner_join((data.frame(
    counts(dds, normalized = TRUE)
  ) %>%
    rownames_to_column("ensembl_gene_id")), by = "ensembl_gene_id")
```

Save normalized counts
```{r}
filename <- "normalized_counts"
write.csv(normalized_counts, paste0(
  "../01_results/deseq2/",
  filename_prefix,
  filename,
  ".csv"
),
row.names = F, quote = F
)

saveRDS(normalized_counts, paste0(
  "../01_results/deseq2/",
  filename_prefix,
  filename,
  ".Rds"
))
```


```{r echo=FALSE}
cat("Saved file :", paste0(
  "../01_results/deseq2/",
  filename_prefix,
  filename,
  ".csv"
))
```

# Supplementary table
```{r}
filename <- "normalized_gene_counts"

excel <- list(
  normalized_counts = normalized_counts
)

write_xlsx(excel, paste0(
  "../04_supplements/",
  filename_prefix,
  filename,
  ".xlsx"
))
```


# R session info
```{r}
utils:::print.sessionInfo(sessionInfo()[-8])
```

# References

- Love MI, Huber W, Anders S. 2014. Moderated estimation of fold change and dispersion for RNA-seq data with DESeq2. Genome Biol 15:550. <a href="https://doi.org/10.1186/s13059-014-0550-8" target="_blank" rel="noopener">Link</a>

<a href="https://bioconductor.org/help/course-materials/2022/CSAMA/lab/2-tuesday/lab-03-rnaseq/rnaseqGene_CSAMA2022.html" target="_blank" rel="noopener">RNA-seq workflow: gene-level exploratory analysis and differential expression</a>

<a href="https://www.bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html" target="_blank" rel="noopener">Analyzing RNA-seq data with DESeq2</a>
